Effects of bariatric surgery loss on resting-state functional connectivity of the reward and default mode network: a pre-registered analysis

### 2021-03-08 manuscript/draft_rsfmri.Rmd
count surgeries

```{r "To do", include=FALSE}
# [x] Upload auf Neurovolt, cf. cobidas reporting guidelines
# [x] Swap mFD with logmFD ??
```

@@

Obesity is a worldwide health issue, entailing huge personal and societal costs. Ex-cess amount of body fat not only affects cardiovascular and metabolic health, but also increases the risk for cognitive decline and dementia later in life (hier eine Ref). Conservative treatment options including behavioral therapy often do not yield the desired weight loss, especially in patients with very high BMI (> 35 kg/m2). Here, bariatric surgery is a viable option to rapidly induce weight loss and improve glycemic status. RYBG is the most invasive, yet most effective surgical pro-cedure, but other techniques like vertical sleeve gastrectomy (VSG) and gadstric banding (GB)are also common [@Berthoud_2011]. 

@@

Precise mechanisms how bariatric surgery leads to altered appetitive and reward signaling are yet to be elucidated. Resting-state functional magnetic resonance imaging (rsfMRI) is a technique to investigate the dynamic organization of the brain. Functional connectivity (FC) networks, i.e. brain regions with correlated neural activity over time, are in anatomical correspondence with specific brain networks involved in cognitive processes, including attention and executive control [@Smith_2009]. The RN (RN), processing hedonic value and internal motivation, and the default mode network (DMN), a higher-order network, involved in interoception and governing shifts between external-internal processes,  are promising candidates to mediate altered central regulation of food intake after bariatric surgery.
The RN (RN), comprises the ventromedial prefrontal cortex (vmPFC), the striatum (STR), the amygdala (AMY) and the anterior insula (antINS) [@Odoherty_2004;@Liu_2011]. These brain regions have been found to guide food valuation processes and decision-making in humans [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018]. Frequently, obesity has been associated with hyperactivation of RN regions during anticipation of (high-caloric) food cues, and in contrast, reduced activation to actual taste of these foods [@Devoto_2018; @Garcia_2014; @Meng_2020; @Stoeckel_2009; though see @Morys_2020]. resting-state rsfMRI studies also shown increased local FC of re-ward network regions, i.e. NAcc, vmPFC, putamen, insula [@Contreras_2017; @Coveleskie_2015;@Hogenkamp_2016], and altered connectivity with salience, homeostatic and sensorimotor networks [@Lips_2014;  @Wijngaarden_2015]. In bariatric surgery patients, connectivity  within the RN (e.g. putamen, OFC) might be normalized by the surgery, however the evidence is limited due to a lack of longitudinal obese control groups  [@Duan_2020; @Wiemerslage_2016]. Possibly, a reconfiguration of the fronto-striatal brain networks could emerge from altered gut signaling , e.g. changes in ghrelin levels, via hypothalamic-striatal projections [@Li_2019; @Karra_2013; but @Zoon_2018] . 

@@

We tested hypothesis 1 by investigating the interaction of bariatric surgery and time on DMN and RN whole-brain FC. We pre-registed two denoising pipelines, and three covariate schemes. For hypothesis 2, we performed a planned analysis of the group-by-time interaction on aggregated, within-network FC, for two and three time points and the same covariate schemes.

@@

dfw <- tidyr::pivot_wider(data = final,
                          id_cols = c("subj.ID","condition","Sex","Age_BL"),
                          names_from = "tp",
                          values_from = c("BMI","meanFD"))
```

@@

**N patients underwent RYGB, N underwent VSG and N GB.**

@@

<!-- MODEL SETUP --> 
#### Confirmatory Analysis
We tested the pre-registered hypothesis of a  time by group interaction:  $FC=group + time + group * time$. 
As pre-registered, we performed the analysis for baseline and followup time points only, in both denoising pipelines and adjusting for no confounders (CA 1), age, sex and average logmFD of both timepoints (CA2) and age,sex, average logmFD and baseline BMI (CA3). 
 In this model, time was represented as factor because we took into account possible non-linear time courses in the increase and decrease of FC over the course of one year, which may occur depending on the phase of weight management [@Olivo_2017]. The resulting factorial design contained one regressor for each time point per group (see supplements for depiction of design matrix).

@@

# Results
<!-- BASELINE CHARACTERISTICS -->
```{r tableBaselineCharacteristics, include=FALSE}
#dfBL <- create_sample_df(group = "all", tp = "BL")
dfBL <- readRDS(file = "../report/dfBL.rds")

# distributions per group
# BMI
p <- ggplot(data=dfBL, aes(group = condition, x=BMI, fill=condition)) +
  geom_density(alpha=0.6) +
  scale_fill_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")) +
  theme_bw()
# log mean FD
p <- ggplot(data=dfBL, aes(group = condition, x=logmFD, fill=condition)) +
  geom_density(alpha=0.6) +
  scale_fill_manual(
    values = c("#00305E55", "#C0004555"),
    labels = c("intervention", "control")) +
  theme_bw()
# Age
p <- ggplot(data=dfBL, aes(group = condition, x=Age_BL, fill=condition)) +
  geom_density(alpha=0.6) +
  scale_fill_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")) +
  theme_bw()

>>>>

```{r "To do", include=FALSE}
# [ ] add grouptime2tp results
# [ ] check stringency
# [ ] check model names
# [ ] add dropout to discussion
# [ ] add comorbidities to methods
# [ ] add imputations to methods
```

@@

Obesity is a worldwide health issue, entailing huge personal and societal costs. Excess amount of body fat not only affects cardiovascular and metabolic health, but also increases the risk for cognitive decline and dementia later in life (hier eine Ref). Conservative treatment options including behavioral therapy often do not yield the desired weight loss, especially in patients with very high BMI ($> 35$ kg/m$^2$). Here, bariatric surgery is a viable option to rapidly induce weight loss and improve glycemic status. RYBG is the most invasive, yet most effective surgical pro-cedure, but other techniques like vertical sleeve gastrectomy (VSG) and gastric banding (GB) are also common [@Berthoud_2011]. 

@@

Precise mechanisms how bariatric surgery leads to altered appetitive and reward signaling are yet to be elucidated. Resting-state functional magnetic resonance imaging (rsfMRI) is a technique to investigate the dynamic organization of the brain. Functional connectivity (FC) networks, i.e. brain regions with correlated neural activity over time, are in anatomical correspondence with specific brain networks involved in cognitive processes, including attention and executive control [@Smith_2009]. The reward network (RN), processing hedonic value and internal motivation, and the default mode network (DMN), a higher-order network, involved in interoception and governing shifts between external-internal processes,  are promising candidates to mediate altered central regulation of food intake after bariatric surgery.
The RN (RN), comprises the ventromedial prefrontal cortex (vmPFC), the striatum, the amygdala and the anterior insula  [@Odoherty_2004;@Liu_2011]. These brain regions have been found to guide food valuation processes and decision-making in humans [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018]. Frequently, obesity has been associated with hyperactivation of RN regions during anticipation of (high-caloric) food cues, and in contrast, reduced activation to actual taste of these foods [@Devoto_2018; @Garcia_2014; @Meng_2020; @Stoeckel_2009; though see @Morys_2020]. resting-state rsfMRI studies also shown increased local FC of re-ward network regions, i.e. NAcc, vmPFC, putamen, insula [@Contreras_2017; @Coveleskie_2015;@Hogenkamp_2016], and altered connectivity with salience, homeostatic and sensorimotor networks [@Lips_2014;  @Wijngaarden_2015]. In bariatric surgery patients, connectivity  within the RN (e.g. putamen, OFC) might be normalized by the surgery, however the evidence is limited due to a lack of longitudinal obese control groups  [@Duan_2020; @Wiemerslage_2016]. Possibly, a reconfiguration of the fronto-striatal brain networks could emerge from altered gut signaling , e.g. changes in ghrelin levels, via hypothalamic-striatal projections [@Li_2019; @Karra_2013; but @Zoon_2018] . 

@@

We tested hypothesis 1 by investigating the interaction of bariatric surgery and time on DMN and RN whole-brain FC. We pre-registered two denoising pipelines, and three covariate schemes. For hypothesis 2, we performed a planned analysis of the group-by-time interaction on aggregated, within-network FC, for two and three time points and the same covariate schemes.

@@

dfw <- tidyr::pivot_wider(data = final,
                          id_cols = c("subj.ID","condition","Sex","Age_BL"),
                          names_from = "tp",
                          values_from = c("BMI","meanFD"))
# patients at baseline
dfBL <- readRDS(file = "../report/dfBL.rds")

# surgeries
data.frame(table(subj_df$interv, subj_df$condition, exclude=NULL))
# why NA ?
subj_df$intervention_de[is.na(subj_df$interv)]
sum(subj_df$comorbidities[subj_df$condition=="IG"])
sum(subj_df$comorbidities[subj_df$condition=="KG"])

subj_df$comorbidities
```

@@

`r nrow(subj_df[subj_df$condition=="IG" & subj_df$interv=="RYGB" & !is.na(subj_df$interv),])` patients underwent RYGB, `r nrow(subj_df[subj_df$condition=="IG" & subj_df$interv=="VSG" & !is.na(subj_df$interv),])` underwent VSG and `r nrow(subj_df[subj_df$condition=="IG" & subj_df$interv=="GB" & !is.na(subj_df$interv),])` GB, for `r nrow(subj_df[subj_df$condition=="IG" & is.na(subj_df$interv),])` patients in the BARS group information was not available.

@@

<!-- MODEL SETUP --> 
#### Confirmatory Analysis
We tested the pre-registered hypothesis of a  time by group interaction:  $FC=group + time + group * time$. 
As pre-registered, we performed the analysis for baseline and followup time points only, in both denoising pipelines and adjusting for no confounders (CA 1), age, sex and average logmFD of both timepoints (CA2) and age, sex, average logmFD and baseline BMI (CA3). 
In this model, time was represented as factor because we took into account possible non-linear time courses in the increase and decrease of FC over the course of one year, which may occur depending on the phase of weight management [@Olivo_2017]. The resulting factorial design contained one regressor for each time point per group (see supplements for depiction of design matrix).

@@

# Results
<!-- BASELINE CHARACTERISTICS -->
```{r tableBaselineCharacteristics, include=FALSE}
# distributions per group
# BMI
p <- ggplot(data=dfBL, aes(group = condition, x=BMI, fill=condition)) +
  geom_density(alpha=0.6) +
  scale_fill_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")) +
  theme_bw()
# log mean FD
p <- ggplot(data=dfBL, aes(group = condition, x=logmFD, fill=condition)) +
  geom_density(alpha=0.6) +
  scale_fill_manual(
    values = c("#00305E55", "#C0004555"),
    labels = c("intervention", "control")) +
  theme_bw()
# Age
p <- ggplot(data=dfBL, aes(group = condition, x=Age_BL, fill=condition)) +
  geom_density(alpha=0.6) +
  scale_fill_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")) +
  theme_bw()
